---
title: "Spotify Audio Features Clustering and Visualization"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---

```{r, include=FALSE}
setwd("~/Development/Python/SpotifyUnsupervised")
require(cluster)
##require(fpc)
require(caret)
require(ggplot2)
require(jsonlite)

# Set seed for reproducibility
set.seed(7)

# Load data
features <- read.csv("audiofeatures.csv")
df_features <- features
info <- read_json('tracksinfo.json', simplifyVector = TRUE)

info <- info$track
uri <- as.data.frame(features$uri)
colnames(uri) <- c('uri')

info <- as.data.frame(cbind(info$name, info$uri, info$album$name, info$artists))
colnames(info) <- c('name', 'uri', 'album_name', 'artists')
```

## Summary of songs
```{r, echo=FALSE}
summary(features)
```

## Feature selection
```{r}
# Remove highly correlated features
correlationMatrix <- cor(features[,1:7])
highlyCorrelated <- findCorrelation(correlationMatrix, cutoff=0.5, verbose=TRUE)
print(highlyCorrelated)
# Feature 1 and 6 are highly correlation, so I removed one of them (1)
features[1] <- NULL
# Remove feature valence because I'm not interested in it
features[6] <- NULL

# This dataframe will be needed later
features_with_uri <- features
features$uri <- NULL
```

## Determine best number of clusters
```{r}
# Calculate the avg silhouette for k = 2,... k = 10
avg.sil <- sapply(2:10, function(x) pam(features, x)$silinfo$avg.width)
avg.sil.df <- data.frame(k = 2:10, silhouette = avg.sil)

ggplot(avg.sil.df, aes(x = k, y = silhouette)) +
  geom_point() +
  geom_line() +
  ggtitle("Average Silhouette for k=2, ..., k=10")

# Calculate the Gap Statistic
c <- clusGap(features, pam, 10, B = 100, verbose = interactive())
plot(c, main = "Gap statistic for k=2, ..., k=10")
```



While 3 clusters seem to be the best parameter, I went for 5 since one of my
main interests is to see how the songs are split between the clusters, and 3 was not
enough.

## The clustering
```{r}
# Perform the cluster
main_clustering <- pam(features, 5, trace.lev = 1)

# Cluster information
print(main_clustering$clusinfo)

# Average silhouette per cluster
print(main_clustering$silinfo$avg.width)
```

```{r}
main_clustering_silhouette <- silhouette(main_clustering)
clusplot(main_clustering, shade=TRUE, color=TRUE)
plot(main_clustering_silhouette, col='black', border='gray',
     main='Silhouette plot of the cluster')
```

The items with negative silhouette are songs that are most probably assigned to the wrong cluster.

```{r}
# Merge the clustering results with the track information
clustering_with_uri <- data.frame(clustering = main_clustering$clustering, uri = uri$uri)
clustering_with_track_info <- merge(clustering_with_uri, info)

# Unlist some values so it is easy to read
clustering_with_track_info$name <- unlist(clustering_with_track_info$name)
clustering_with_track_info$album_name <- unlist(clustering_with_track_info$album_name)

clustering_with_track_info$artists <- as.character(clustering_with_track_info$artists)
```

```{r}
# CLUSTER 4
# Show the top of the cluster 4
head(clustering_with_track_info[clustering_with_track_info$clustering == 4,])
clustering_with_track_info$album_name <- as.factor(clustering_with_track_info$album_name)
```

The content of this cluster is mostly made of songs from videogames (Child of Light, Limbo, and Transistor), songs from movie soundtracks (Interstellar, Lord of The Rings, Mad Max), and other highly instrumental songs; none of the first 40 songs have vocals.

```{r}
head(clustering_with_track_info[clustering_with_track_info$clustering == 3,])
summary(clustering_with_track_info[clustering_with_track_info$clustering == 3,])
```

This cluster shows more diversity than the previous one. It is mostly made of songs
from both Above & Beyond's Acoustic albums, Kanye West music, and songs from Darren Korb, composer of the soundtrack of the Transistor and Bastion videogames.

What happen if we perform a new clustering with the content of this cluster?

```{r}
# Cluster the content of cluster 3
c3_track_info <- clustering_with_track_info[clustering_with_track_info$clustering == 3,]
c3 <- clustering_with_uri[clustering_with_uri$clustering == 3,]
c3 <- merge(c3, features_with_uri, by = 'uri')

c3_features <- c3[c('liveness', 'speechiness', 'acousticness', 'instrumentalness',
                    'danceability')]
c3$clustering <- NULL
c3_clustering <- pam(c3_features, 4, trace.lev = 1)
print(c3_clustering$clusinfo)

c3_clustering_sil <- silhouette(c3_clustering)
clusplot(c3_clustering, shade=TRUE, color=TRUE)
plot(c3_clustering_sil, col='black', border='gray',
     main='Silhouette plot of the cluster')
```

```{r}
## These are the songs of the 2 cluster of the clustering done on cluster 3
c3_results <- data.frame(cluster = c3_clustering$clustering, uri = c3_track_info$uri,
                         album = c3_track_info$album_name, artist = c3_track_info$artists)


cluster.3 <- data.frame(c3_features, clustering = c3_clustering$clustering)
write.csv(x = cluster.3, file="cluster_3.csv", row.names = FALSE)
```
